A Process for Emulating Comparative Oncology Trials with Real-world Evidence Studies

### 2025-03-04 manuscript/manuscript.tex
Manuscript updated

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={A Process for the Emulation of Comparative Oncology Trials with Real-world Evidence (ENCORE)},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{A Process for the Emulation of Comparative Oncology Trials with
Real-world Evidence (ENCORE)}
\author{}
\date{}

@@

\textbf{Authors}: Janick Weberpals\textsuperscript{1}, Kenneth L.
Kehl\textsuperscript{2}, Donna R. Rivera\textsuperscript{3}, Pallavi
Mishra-Kalyani\textsuperscript{3}, Georg Hahn\textsuperscript{1},
Priyanka Anand\textsuperscript{1}, Yanina Natanzon\textsuperscript{4},
Andrew J. Belli\textsuperscript{5}, Ching-Kun Wang\textsuperscript{5},
Jenna Collins\textsuperscript{6}\textbf{,}
JonathanKish\textsuperscript{6}\textbf{,} Janet
Espirito\textsuperscript{7}, Nicholas J Robert\textsuperscript{7},
Robert J. Glynn\textsuperscript{1} ,Sebastian
Schneeweiss\textsuperscript{1}, Shirley V. Wang\textsuperscript{1}

@@

\ul{\textbf{Competing Interests Statement:}} Dr.~Weberpals is now an
employee of AstraZeneca and owns stocks in AstraZeneca. Dr.~Kehl has
received research funding from Meta, Inc.~to his institution. Drs.
Espirito and Robert are employees of McKesson and own McKesson stock.
Dr.~Wang has consulted ad hoc for Exponent Inc.~and MITRE a federally
funded research center for the Centers for Medicare and Medicaid
Services on unrelated work. Dr.~Schneeweiss is participating in
investigator-initiated grants to the Brigham and Women's Hospital from
Boehringer Ingelheim, Takeda, and UCB unrelated to the topic of this
study. He owns equity in Aetion Inc., a software manufacturer. He is an
advisor to Temedica GmbH, a patient-oriented data generation company.
His interests were declared, reviewed, and approved by the Brigham and
Women's Hospital in accordance with their institutional compliance
policies.

@@

\emph{Manuscript last updated: 2025-02-26 21:44:25.38931}

@@

Randomized controlled trials (RCTs) have been the gold standard for
establishing the efficacy and safety of medical products. With the
advent of the 21\textsuperscript{st} Century Cures Act
directive,\textsuperscript{1} the Food and Drug Administration (FDA)
established a framework to increasingly consider real-world evidence
(RWE) generated from routine-care health data such as electronic health
records (EHR) to evaluate and contextualize the comparative safety and
effectiveness of novel cancer therapies.\textsuperscript{2} Accounting
for 21\% of all approvals, oncology was the disease area with the most
FDA drug approvals in 2023.\textsuperscript{3} RWE has particularly
important potential to complement evidence coming from RCTs in the field
of precision oncology, where potential use cases include the assessment
of effectiveness in rare patient populations that are underrepresented
in RCTs, the construction of external control arms in single-arm trials
where active recruitment to a randomized trial may not be feasible, or
the discovery of biomarkers among pan-tumor populations that harbor
specific genomic and immuno-pathological signatures.

However, the validity and transportability of results derived between
RWE studies and RCTs can depend on many factors. Frequently referenced
limitations include missing data, small sample sizes, data
discontinuity,\textsuperscript{4, 5} adoption of changes in guidelines
in real-world care and the inability to measure and emulate common
eligibility criteria, including prognostic factors, and standardized
response assessments in real-world data (RWD).\textsuperscript{6} While
there are already published examples of oncology trial
emulations,\textsuperscript{6--8} a systematic and scaled approach to
emulate a diverse set of different oncology trials with multiple
heterogeneous databases is necessary to gain confidence in the validity
of RWE studies and to provide context as to which questions can be
validly answered.

The RCT DUPLICATE initiative\textsuperscript{9} increased our
understanding of when RWE studies can come to causal conclusions on
treatment effects by benchmarking results against RCTs under the
assumption that each RCT finding reflects a causal treatment effect. In
settings where the RCT designs could be emulated well, RWE studies came
to the same conclusions.\textsuperscript{10} However, prior work from
RCT-DUPLICATE has focused primarily on emulating trials in the
cardio-metabolic, renal, and pulmonary clinical areas using claims
databases.

The \emph{Emulation of Comparative Oncology Trials with Real-world
Evidence} (ENCORE) project\textsuperscript{11} aims to extend this work
to the field of oncology. Studies in oncology come with their own unique
set of challenges which must be systematically explored and understood.
Building on a process co-developed with the FDA through RCT
DUPLICATE,\textsuperscript{9} this expansion to oncology will emulate 12
randomized clinical trials using multiple specialty oncology EHR data
sources. The process will emphasize transparency and include documented
data fitness assessment\textsuperscript{12, 13} for each RWD source with
respect to each trial emulation as well as extensive sensitivity
analyses to assess robustness of findings.

@@

The focus of ENCORE is to maximize potential learnings on when RWE
studies can or cannot yield similar results compared to RCTs. To that
end, the emphasis of the project is on trials of therapies for the most
common cancers for which there has been substantial therapeutic
development in recent years. After careful review and exchange with
clinical and regulatory experts, four cancer indications were
identified: non-small cell lung cancer, breast cancer, colorectal cancer
and multiple myeloma. For each cancer type, we aim to conduct three
trial emulations using multiple databases accessible for the scope of
this project (i.e., the total number of emulations will equal 12 trials
x \emph{n} databases which are found fit-for-purpose for each trial).

@@

The rationale and operationalization of each criterion is listed in
detail in Table~\ref{tbl-criteria}. We will mainly consider pivotal
interventional, randomized trials after 2011 because treatment
guidelines among included cancer indications have undergone significant
changes in recent years. Due to the rapid adoption of new breakthrough
therapies in routine care, it is unlikely to find patients who may be
still treated with outdated treatment regimens in current clinical
practice. We also did not include trials with results published too
recently. This decision was in order to allow for enough data and
follow-up time accrual in the databases used for this project. Although
there have been substantial methodological advancements to increase our
understanding on the emulation and comparison of real-world
progression-free survival (PFS) and objective response rates (ORR) to a
RECISTv1.1\textsuperscript{15}-based PFS and ORR assessment in
RCTs\textsuperscript{16, 17}, imaging-based evaluations still hold a
level of granularity which may not be necessarily reflected in
chart-abstracted assessments of a patient's progression in routine
care.\textsuperscript{18, 19} In addition, the timing and cadence of
intervals between progression assessments can differ between RCTs and
routine care which may result in measurement error and
bias.\textsuperscript{20} Given this, and the large number of other
methodological challenges like missing data, small sample sizes, data
discontinuity and rapidly changing guideline treatments, we focus on the
emulation of overall survival (OS) as the endpoint of interest.
Therefore, we include only trials that have reported overall survival
(OS) as one of the pre-specified endpoints in the protocol.

While most trial eligibility criteria could be operationalized in an
automated fashion, our final criterion was an assessment of emulation
feasibility and clinical relevance. This criterion involved extensive
human review. The critical points considered in this step include an
initial feasibility assessment of the data fitness,\textsuperscript{19}
including assessment of whether critical eligibility criteria (e.g.,
biomarker status) and prognostic factors (e.g., ECOG performance score)
are measurable and whether preliminary sample size counts are
reasonable. Lastly, trial candidates were ranked and shortlisted into
primary and runner-up candidates based on their clinical and regulatory
relevance.

A list of tentative, shortlisted primary candidates is presented in
Table~\ref{tbl-rcts} and the corresponding selection process is
illustrated in the CONSORT diagrams (Supplementary Figures 1-4).
Naturally, the majority of trials cover advanced (locally advanced,
inoperable, recurrent/progressive disease) or metastatic cancer
populations because a large proportion of drug development efforts start
in these settings. A key objective that we aim to foster with the
shortlisted trials is to achieve a better understanding how different
disease settings (early, late), line settings ({[}neo{]}adjuvant, first
line, advanced lines of therapy), therapy protocols (monotherapy,
combination therapy) and population characteristics (simple versus
complex genetic or immunological signatures) can be emulated using RWD.
If more thorough feasibility assessments suggest that the threat of bias
from mis-measurement of key study parameters or residual confounding
remains high, or that the sample size is not sufficient, then runner-up
candidates will be considered instead.

@@

For each shortlisted and selected RCT, a detailed protocol,
pre-specifying key elements of the trial emulation, will be developed
using the HARPER protocol template\textsuperscript{22} recommended for
regulatory submissions of RWE studies\textsuperscript{23} and will be
registered on ClinicalTrials.gov after careful review by a clinical and
FDA regulatory expert panel. Following the target trial emulation
framework, we will provide an explicit statement and rationale on how
each element will be emulated including database selection, covariate
measurement, operationalization of key eligibility criteria, study
design, data analysis and causal contrasts of
interest.\textsuperscript{24, 25} Since it is common that oncology RCTs
update OS estimates periodically based on accrued follow-up time, the
protocol will give a brief summary of each emulated RCT and specify
which target OS estimates will be used to compare agreement metrics to
(see Section~\ref{sec-agreement-metrics}). All eligibility criteria will
be extracted based on publications, publicly available protocols and
statistical analysis plans of the selected RCT.

@@

We will additionally provide a definition on how exposures, outcomes and
covariates are exactly defined and operationalized in each respective
database. There will be special emphasis on how exposure, in context of
their respective disease and line of therapy settings, and the OS
outcome will be emulated. For all considered databases, the OS endpoint
is typically a composite that is derived from different sources
comprising EHR documentation, social security death index, obituary and
other linkages. Given that not all relevant sources that provide
mortality data are synchronized and updated uniformly, sensitivity
analyses with more conservative (i.e., earlier) censoring dates will be
considered for each trial emulation to mitigate the potential impact of
ghost-time bias.\textsuperscript{26}

@@

\paragraph{\texorpdfstring{\textbf{Statistical
power}}{Statistical power}}\label{statistical-power}

@@

To establish an analytic cohort, key eligibility criteria will be
applied in which patients with missing values in eligibility criteria
are considered eligible in the respective attrition steps to allow for
thorough missing data investigations. These missing data investigations
will empirically assess assumptions on potential underlying missingness
mechanisms according to Rubin's classification of missing data (i.e.,
missing completely at random {[}MCAR{]}, missing at random {[}MAR{]} and
missing not at random {[}MNAR{]}).\textsuperscript{32} To that end, we
will adopt a principled process on missing data that was developed as
part of a FDA Sentinel Innovation Center causal inference workstream
that empirically evaluates different aspects across partially observed
covariates based on three group diagnostics.\textsuperscript{33, 34} In
brief, these diagnostics cover (1) comparisons of patients
characteristics with and without an observed level of the partially
observed covariate, (2) ability to predict missingness given observed
data, and (3) assessments if outcomes between patients with a missing
value are systematically different. Together with expert domain
knowledge and assumptions about the underlying missing data structure
through canonical causal diagrams,\textsuperscript{35} this will inform
decisions regarding the in- or exclusion of patients with missing values
in key eligibility criteria and potential sensitivity analyses to assess
the robustness of these decisions.\textsuperscript{36}

@@

Due to its ubiquity in oncology trials, the primary parameter of
interest in ENCORE will be defined as the marginal hazard ratio (HR)
coefficient for the treatment comparison for time to all-cause mortality
(OS).\textsuperscript{39} However, the HR has many limitations that can
make the comparison of results across varying follow-up times between
RCT and emulations challenging, including its non-collapsibility,
in-built selection bias and the requirement of proportional
hazards.\textsuperscript{40, 41} For this reason, we will also consider
alternative endpoints on an absolute risk scale such as median survival
times, survival probabilities at pre-defined time points during
follow-up\textsuperscript{42} and restricted mean survival times as
secondary endpoints of interest.

For the estimation of marginal treatment effects, we will employ
propensity score methods to adjust for measured confounding between
treatment arms. The selection of important prognostic covariates will be
based on expert clinical knowledge and published literature on
prognostic scores in oncology.\textsuperscript{43} The implementation of
propensity scores in combination with multiple imputation will follow
the `\emph{within}' methodology as described by Leyrat et
al.\textsuperscript{44, 45} That is, propensity score matching or
weighting will be applied to each imputed dataset. The marginal
treatment effect will then be estimated in each imputed and matched or
weighted dataset separately and pooled into a final estimate following
Rubin's rule.\textsuperscript{46, 47} This approach has been shown to
lead to unbiased estimates across different simulated scenarios with a
sufficient estimation of the variance.\textsuperscript{44}

To asses the balance of pre-exposure covariates after matching or
weighting on each imputed dataset, the average SMD and corresponding
minimum and maximum SMD range will be visualized (see example in
Figure~\ref{fig-balance}). Covariate balance is typically considered at
a SMD \textless{} 0.1.\textsuperscript{48} Further, we will compute the
average post-matching or post-weighting
C-statistics.\textsuperscript{49} In addition, we will use a published
prognostic score for OS\textsuperscript{43} as a balance measure to
visually assess if the prognostic score is balanced between treatment
arms after propensity score matching or weighting as this approach was
described to show the highest correlations with bias compared with other
balance measures and not affected by model
misspecification.\textsuperscript{50}

Similarly, survival probabilities for individual time points will be
estimated in each imputed and propensity score matched or weighted
dataset according to the Kaplan-Meier method.\textsuperscript{42} Since
survival probabilities typically do not follow normal distributions
which are required to apply Rubin's rule, these will be transformed
through a complementary log-log transformation \(log(-log(1-pr(surv)))\)
with \(pr(surv)\) denoting the survival probability at a given time
during follow-up.\textsuperscript{51, 52} The transformed survival
probabilities are then pooled across imputed datasets and individual
time points following Rubin's rule and back-transformed via
\(1-exp(-exp(qbar))\) with \(qbar\) denoting the pooled survival
probability. The median survival time can be finally determined by
extracting the time point during follow-up at which the survival
probability drops below 0.5 for the first time.

@@

Building on an established approach based on the RCT DUPLICATE
project\textsuperscript{9, 30}, the ENCORE project aims to emulate 12
oncology trials using multiple EHR data sources to inform the potential
use of RWE for regulatory science in oncology. The project focuses on
four common cancers and will evaluate the agreement of treatment effect
estimates between RCTs and their respective emulations. Historically,
administrative health claims databases have been the backbone of most
research in RWE. With increasing access to EHR data and a maturing set
of methodological approaches to draw causal inferences from such data,
the ENCORE project is one of the first of its kind to evaluate when and
how RWD can be used to deliver similar causal conclusions compared to
RCTs in the field of oncology.

The RCT DUPLICATE project\textsuperscript{9, 30} identified multiple
emulation challenges which we expect to also encounter in ENCORE. One
particular aspect that we may not always be able to emulate is the exact
distribution of patient characteristics of the trial population. This
can be due to the lack of data granularity and comprehensiveness to
emulate relevant eligibility criteria or the fact that large pivotal
trials in oncology are typically conducted in multiple countries
worldwide, which we will not be able to mirror given that all considered
databases reflect the US only. This may be a critical factor especially
given that the pathophysiology, prognosis and factors that drive
heterogeneous treatment outcomes of certain cancers differ between
countries. For example, some cancer indications (e.g., GI cancers) or
some genetic mutations (e.g., EGFR mutations in lung adenocarcinoma) are
much more prevalent in Asian countries compared to the US or Europe.

Another common challenge in the emulation of oncology trials is the
estimation of an ``intention-to-treat'' (ITT) analogous estimand which
is usually the primary estimand reported in oncology RCTs. Due to
intercurrent events, such as non-adherence, crossover of a high
proportion of patients from the control to the intervention arm, or
differences in subsequent therapy lines there may be estimand
differences between trial and emulation.\textsuperscript{53} Although
this is also a common challenge in the analysis of
RCTs,\textsuperscript{27} treatment protocols in routine care are often
observed to be less stringent compared to trials.\textsuperscript{8} In
order to derive comparable estimands it is therefore crucial to
understand and contextualize the proportion and timing of treatment
switching and discontinuation in both the trial and its emulation.

This issue may be even augmented with more complex treatment protocols
like combination regimens (e.g., CheckMate9LA) as compared to
monotherapies (e.g., CheckMate057) since the ascertainment of the
exposure needs to happen over a pre-defined time window which is often
difficult to calibrate in RWD. As a result, exposure misclassification
(due to overly ascertainment windows) and selection bias (due to overly
long ascertainment windows) are common trade-offs in such scenarios.
Alternative analytic approaches which target a per-protocol estimand,
such as the clone-censor-weight design,\textsuperscript{54} may be
viable options if the ITT estimand cannot be estimated due to the
aforementioned parameters, provided that necessary covariate
measurements are available to account for the artificial censoring
introduced with these methods.

@@

\bibitem[\citeproctext]{ref-hernan2010hazards}
\textbf{40}. Hernán MA: The hazards of hazard ratios. Epidemiology
21:13--15, 2010

\bibitem[\citeproctext]{ref-stensrud2020test}
\textbf{41}. Stensrud MJ, Hernán MA:
\href{https://doi.org/10.1001/jama.2020.1267}{Why test for proportional
hazards?} Jama 323:1401--1402, 2020

\bibitem[\citeproctext]{ref-kaplan1958nonparametric}
\textbf{42}. Kaplan EL, Meier P: Nonparametric estimation from
incomplete observations. Journal of the American statistical association
53:457--481, 1958

\bibitem[\citeproctext]{ref-becker2020enhanced}
\textbf{43}. Becker T, Weberpals J, Jegg A, et al: An enhanced
prognostic score for overall survival of patients with cancer derived
from a large real-world cohort. Annals of Oncology 31:1561--1568, 2020

\bibitem[\citeproctext]{ref-leyrat2019propensity}
\textbf{44}. Leyrat C, Seaman SR, White IR, et al: Propensity score
analysis with partially observed covariates: How should multiple
imputation be used? Statistical methods in medical research 28:3--19,
2019

\bibitem[\citeproctext]{ref-pishgar2020matchthem}
\textbf{45}. Pishgar F, Greifer N, Leyrat C, et al: MatchThem:: Matching
and weighting after multiple imputation. arXiv preprint arXiv:200911772,
2020

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={A Process for Emulating Comparative Oncology Trials with Real-world Evidence Studies},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{A Process for Emulating Comparative Oncology Trials with
Real-world Evidence Studies}
\author{}
\date{}

@@

\textbf{Authors}: Janick Weberpals\textsuperscript{1}, Kenneth L.
Kehl\textsuperscript{2}, Donna R. Rivera\textsuperscript{3}, Pallavi
Mishra-Kalyani\textsuperscript{3}, Georg Hahn\textsuperscript{1},
Priyanka Anand\textsuperscript{1}, Yanina Natanzon\textsuperscript{4},
Andrew J. Belli\textsuperscript{5}, Ching-Kun Wang\textsuperscript{5},
Jenna Collins\textsuperscript{6}\textbf{,}
JonathanKish\textsuperscript{6}\textbf{,} Janet
Espirito\textsuperscript{7}, Nicholas J Robert\textsuperscript{7},
Robert J. Glynn\textsuperscript{1}, Sebastian
Schneeweiss\textsuperscript{1}, Shirley V. Wang\textsuperscript{1}

@@

\ul{\textbf{Competing Interests Statement:}} Dr.~Weberpals is now an
employee of AstraZeneca and owns stocks in AstraZeneca. Dr.~Kehl has
received research funding from Meta, Inc.~to his institution. Drs.
Espirito and Robert are employees of McKesson and own McKesson stock.
Dr.~Wang has consulted ad hoc for Exponent Inc.~and MITRE a federally
funded research center for the Centers for Medicare and Medicaid
Services on unrelated work. Dr.~Schneeweiss is participating in
investigator-initiated grants to the Brigham and Women's Hospital from
Bayer and UCB unrelated to the topic of this study. He consults for and
owns equity in Aetion Inc., a software manufacturer. He is an advisor to
Temedica GmbH, a patient-oriented data generation company. His interests
were declared, reviewed, and approved by the Brigham and Women's
Hospital in accordance with their institutional compliance policies.

@@

\emph{Manuscript last updated: 2025-03-04 18:43:42.372054}

@@

Randomized controlled trials (RCTs) have been the go-to methodology for
establishing the efficacy and safety of medical products. Under the
21\textsuperscript{st} Century Cures Act directive,\textsuperscript{1}
the Food and Drug Administration (FDA) established a framework to
increasingly consider real-world evidence (RWE) generated from
routine-care health data such as electronic health records (EHR) to
evaluate and contextualize the comparative safety and effectiveness of
novel cancer therapies.\textsuperscript{2} Accounting for 21\% of all
approvals, oncology was the disease area with the most FDA drug
approvals in 2023.\textsuperscript{3} RWE has particularly important
potential to complement evidence coming from RCTs in the field of
precision oncology, where potential use cases include the assessment of
effectiveness in specific patient populations that are underrepresented
in RCTs, the construction of external control arms in single-arm trials
where active recruitment to a randomized trial may not be feasible, or
the discovery of biomarkers among pan-tumor populations that harbor
specific genomic and immuno-pathological signatures.

However, the validity and transportability of results derived between
RWE studies and RCTs depends on multiple factors. Frequently referenced
limitations include, lack of baseline randomization, missing data, small
study sizes, data discontinuity,\textsuperscript{4, 5} adoption of
changes in guidelines in real-world care and the inability to measure
and emulate common eligibility criteria, including prognostic factors,
and standardized response assessments in real-world data
(RWD).\textsuperscript{6} While there are already published examples of
oncology trial emulations,\textsuperscript{6--8} a systematic and scaled
approach to emulate a diverse set of different oncology trials with
multiple heterogeneous databases is necessary to gain confidence in the
validity of RWE studies and to provide context as to which questions can
be validly answered.

The RCT DUPLICATE initiative\textsuperscript{9} increased our
understanding of when RWE studies can come to causal conclusions on
treatment effects by benchmarking results against RCTs under the
assumption that each RCT finding reflects a causal treatment effect. In
settings where the RCT designs could be emulated well, RWE studies came
to the same conclusions.\textsuperscript{10} However, prior work from
RCT-DUPLICATE has focused primarily on emulating trials in non-oncology
settings using claims databases.

The \emph{Emulation of Comparative Oncology Trials with Real-world
Evidence} (ENCORE) project\textsuperscript{11} aims to extend this work
to oncology. Studies in oncology come with their own unique set of
challenges which must be systematically explored and understood.
Building on a process co-developed with the FDA through RCT
DUPLICATE,\textsuperscript{9} this expansion to oncology will emulate 12
randomized clinical trials using multiple specialty oncology EHR data
sources. The process will emphasize transparency and include documented
data fitness assessment\textsuperscript{12, 13} for each RWD source with
respect to each trial emulation as well as extensive sensitivity
analyses to assess robustness of findings.

@@

The focus of ENCORE is to maximize potential learnings on when RWE
studies can or cannot yield similar results compared to RCTs. To that
end, the emphasis of the project is on trials of therapies for the most
common cancers for which there has been substantial therapeutic
development in recent years. After review and discussion with clinical
and regulatory experts, four cancer indications were identified:
non-small cell lung cancer, breast cancer, colorectal cancer and
multiple myeloma. For each cancer type, we aim to conduct three trial
emulations using multiple databases accessible for the scope of this
project (i.e., the total number of emulations will equal 12 trials x
\emph{n} databases which are found fit-for-purpose for each trial).

@@

The rationale and operationalization of each criterion is listed in
detail in Table~\ref{tbl-criteria}. We will mainly consider pivotal
interventional, randomized trials after 2011 because treatment
guidelines among included cancer indications have undergone significant
changes in recent years. Due to the rapid adoption of new breakthrough
therapies in routine care, it is unlikely to find patients who may be
still treated with outdated treatment regimens in current clinical
practice. Conversely, we excluded trials with results published too
recently. To allow for enough data and follow-up time accrual in the
databases used for this project. Although there have been substantial
methodological advancements to increase our understanding on the
emulation and comparison of real-world progression-free survival (PFS)
and objective response rates (ORR) to a
RECISTv1.1\textsuperscript{15}-based PFS and ORR assessment in
RCTs\textsuperscript{16, 17}, imaging-based evaluations still hold a
level of granularity which may not be necessarily reflected in
chart-abstracted assessments of a patient's progression in routine
care.\textsuperscript{18, 19} The timing and cadence of intervals
between progression assessments can differ between RCTs and routine care
which may result in measurement error and bias.\textsuperscript{20}
Given this, and the large number of other methodological challenges like
missing data, small sample sizes, data discontinuity and rapidly
changing guideline treatments, we focus on the emulation of overall
survival (OS) as the endpoint of interest. Therefore, we include only
trials that have reported overall survival (OS) as one of the
pre-specified endpoints in the protocol.

While most trial eligibility criteria could be operationalized in an
automated fashion, our final criterion was an assessment of emulation
feasibility and clinical relevance. This criterion involved extensive
human review. The critical points considered in this step include an
initial feasibility assessment of the data fitness,\textsuperscript{19}
including assessment of whether critical eligibility criteria (e.g.,
biomarker status) and prognostic factors (e.g., ECOG performance score)
are measurable and whether preliminary study size counts are reasonable.
Lastly, trial candidates were ranked and shortlisted into primary and
runner-up candidates based on their clinical and regulatory relevance.

A list of tentative, shortlisted primary candidates is presented in
Table~\ref{tbl-rcts} and the corresponding selection process is
illustrated in the CONSORT diagrams (Supplementary Figures 1-4).
Naturally, the majority of trials cover advanced (locally advanced,
inoperable, recurrent/progressive disease) or metastatic cancer
populations because a large proportion of drug development efforts start
in these settings. A key objective that we aim to foster with the
shortlisted trials is to achieve a better understanding how different
disease settings (early, late), line settings ({[}neo{]}adjuvant, first
line, advanced lines of therapy), therapy protocols (monotherapy,
combination therapy) and population characteristics (simple versus
complex genetic or immunological signatures) can be emulated using RWD.
If more thorough feasibility assessments suggest that the threat of bias
from mis-measurement of key study parameters or residual confounding
remains high, or that the study size is not sufficient, then runner-up
candidates will be considered instead.

@@

For each shortlisted and selected RCT, a detailed protocol,
pre-specifying key elements of the trial emulation, will be developed
using the HARPER protocol template\textsuperscript{22} recommended for
regulatory submissions of RWE studies\textsuperscript{23} and will be
registered on ClinicalTrials.gov after review by a clinical and FDA
regulatory expert panel. Following the target trial emulation framework,
we will provide an explicit statement and rationale on how each element
will be emulated including database selection, covariate measurement,
operationalization of key eligibility criteria, study design, data
analysis and causal contrasts of interest.\textsuperscript{24, 25} Since
it is common that oncology RCTs update OS estimates periodically based
on accrued follow-up time, the protocol will give a brief summary of
each emulated RCT and specify which target OS estimates will be used to
compare agreement metrics to (see Section~\ref{sec-agreement-metrics}).
All eligibility criteria will be extracted based on publications,
publicly available protocols and statistical analysis plans of the
selected RCT.

@@

We will additionally provide a definition on how exposures, outcomes and
covariates are exactly defined and operationalized in each respective
database. There will be special emphasis on how exposure, in context of
their respective disease and line of therapy settings, and the OS
outcome will be emulated. For all considered databases, the OS endpoint
is typically a composite that, depending on the underlying database, can
be derived from different sources comprising EHR documentation, social
security death index, obituary and other linkages. Given that not all
relevant sources that provide mortality data are synchronized and
updated uniformly, sensitivity analyses with more conservative (i.e.,
earlier) censoring dates will be considered for each trial emulation to
mitigate the potential impact of ghost-time bias.\textsuperscript{26}

@@

\paragraph{\texorpdfstring{\textbf{Statistical power
considerations}}{Statistical power considerations}}\label{statistical-power-considerations}

@@

To establish an analytic cohort, key eligibility criteria will be
applied in which patients with missing values in eligibility criteria
are considered eligible in the respective attrition steps to allow for
thorough missing data investigations. These missing data investigations
will empirically assess assumptions on potential underlying missingness
mechanisms according to Rubin's classification of missing data (i.e.,
missing completely at random {[}MCAR{]}, missing at random {[}MAR{]} and
missing not at random {[}MNAR{]}).\textsuperscript{32} To that end, we
will adopt a principled process on missing data that empirically
evaluates different aspects across partially observed covariates based
on three group diagnostics.\textsuperscript{33, 34} These diagnostics
cover (1) comparisons of patients characteristics with and without an
observed level of the partially observed covariate, (2) ability to
predict missingness given observed data, and (3) assessments if outcomes
between patients with a missing value are systematically different.
Together with expert domain knowledge and assumptions about the
underlying missing data structure through canonical causal
diagrams,\textsuperscript{35} this will inform decisions regarding the
in- or exclusion of patients with missing values in key eligibility
criteria and potential sensitivity analyses to assess the robustness of
these decisions.\textsuperscript{36}

@@

Due to its ubiquity in oncology trials, the primary parameter of
interest in ENCORE will be defined as the marginal hazard ratio (HR)
coefficient for the treatment comparison for time to all-cause mortality
(OS).\textsuperscript{39} We will also consider alternative endpoints on
an absolute risk scale such as median survival times, survival
probabilities at pre-defined time points during
follow-up\textsuperscript{40} and restricted mean survival times as
secondary endpoints of interest.

For the estimation of marginal treatment effects, we will employ
propensity score methods to adjust for measured confounding between
treatment arms. The selection of important prognostic covariates will be
based on expert clinical knowledge and published literature on
prognostic scores in oncology.\textsuperscript{41} The implementation of
propensity scores in combination with multiple imputation will follow
the `\emph{within}' methodology as described by Leyrat et
al.\textsuperscript{42, 43} That is, propensity score matching or
weighting will be applied to each imputed dataset. The marginal
treatment effect will then be estimated in each imputed and matched or
weighted dataset separately and pooled into a final estimate following
Rubin's rule.\textsuperscript{44, 45} This approach has been shown to
lead to unbiased estimates across different simulated scenarios with a
sufficient estimation of the variance.\textsuperscript{42}

To asses the balance of pre-exposure covariates after matching or
weighting on each imputed dataset, the average SMD and corresponding
minimum and maximum SMD range will be visualized (see example in
Figure~\ref{fig-balance}). Covariate balance is typically considered at
a SMD \textless{} 0.1.\textsuperscript{46} Further, we will compute the
average post-matching or post-weighting
C-statistics.\textsuperscript{47} In addition, we will use a published
prognostic score for OS\textsuperscript{41} as a balance measure to
visually assess if the prognostic score is balanced between treatment
arms after propensity score matching or weighting as this approach was
described to show the highest correlations with bias compared with other
balance measures and not affected by model
misspecification.\textsuperscript{48}

Similarly, survival probabilities for individual time points will be
estimated in each imputed and propensity score matched or weighted
dataset according to the Kaplan-Meier method.\textsuperscript{40} Since
survival probabilities typically do not follow normal distributions
which are required to apply Rubin's rule, these will be transformed
through a complementary log-log transformation \(log(-log(1-pr(surv)))\)
with \(pr(surv)\) denoting the survival probability at a given time
during follow-up.\textsuperscript{49, 50} The transformed survival
probabilities are then pooled across imputed datasets and individual
time points following Rubin's rule and back-transformed via
\(1-exp(-exp(qbar))\) with \(qbar\) denoting the pooled survival
probability. The median survival time can be finally determined by
extracting the time point during follow-up at which the survival
probability drops below 0.5 for the first time.

@@

Building on an established approach developed by the RCT DUPLICATE
project\textsuperscript{9, 30}, the ENCORE project aims to emulate 12
oncology trials using four oncology EHR data sources to inform the
potential use of RWE for regulatory science in oncology. The project
focuses on four common cancers and will evaluate the agreement of
treatment effect estimates between RCTs and their respective emulations.
Historically, administrative health claims databases have been the
backbone of most research in RWE. With increasing access to EHR data and
a maturing set of methodological approaches to draw causal inferences
from such data, the ENCORE project is one of the first of its kind to
evaluate when and how RWD can be used to deliver similar causal
conclusions compared to RCTs in the field of oncology.

The RCT DUPLICATE initiative\textsuperscript{9, 30} identified multiple
emulation challenges which we expect to also encounter in ENCORE. One
particular aspect that we may not always be able to emulate is the exact
distribution of patient characteristics of the trial population. This
can be due to the lack of data granularity and comprehensiveness to
emulate relevant eligibility criteria or the fact that large pivotal
trials in oncology are typically conducted in multiple countries
worldwide, which we will not be able to mirror given that all considered
databases reflect the US only. This may be a critical factor especially
given that the pathophysiology, prognosis and factors that drive
heterogeneous treatment outcomes of certain cancers differ between
countries. For example, some cancer indications (e.g., GI cancers) or
some genetic mutations (e.g., EGFR mutations in lung adenocarcinoma) are
much more prevalent in Asian countries compared to the US or Europe.

Another common challenge in the emulation of oncology trials is the
estimation of an ``intention-to-treat'' (ITT) analogous estimand which
is usually the primary estimand reported in oncology RCTs. Due to
intercurrent events, such as non-adherence, crossover of a high
proportion of patients from the control to the intervention arm, or
differences in subsequent therapy lines there may be estimand
differences between trial and emulation.\textsuperscript{51} Although
this is also a common challenge in the analysis of
RCTs,\textsuperscript{27} treatment protocols in routine care are often
observed to be less stringent compared to trials.\textsuperscript{8} In
order to derive comparable estimands it is therefore crucial to
understand and contextualize the proportion and timing of treatment
switching and discontinuation in both the trial and its emulation.

This issue may be even augmented with more complex treatment protocols
like combination regimens (e.g., CheckMate9LA) as compared to
monotherapies (e.g., CheckMate057) since the ascertainment of the
exposure needs to happen over a pre-defined time window which is often
difficult to calibrate in RWD. As a result, exposure misclassification
(due to overly ascertainment windows) and selection bias (due to overly
long ascertainment windows) are common trade-offs in such scenarios.
Alternative analytic approaches which target a per-protocol estimand,
such as the clone-censor-weight design,\textsuperscript{52} may be
viable options if the ITT estimand cannot be estimated due to the
aforementioned parameters, provided that necessary covariate
measurements are available to account for the artificial censoring
introduced with these methods.

@@

\bibitem[\citeproctext]{ref-kaplan1958nonparametric}
\textbf{40}. Kaplan EL, Meier P: Nonparametric estimation from
incomplete observations. Journal of the American statistical association
53:457--481, 1958

\bibitem[\citeproctext]{ref-becker2020enhanced}
\textbf{41}. Becker T, Weberpals J, Jegg A, et al: An enhanced
prognostic score for overall survival of patients with cancer derived
from a large real-world cohort. Annals of Oncology 31:1561--1568, 2020

\bibitem[\citeproctext]{ref-leyrat2019propensity}
\textbf{42}. Leyrat C, Seaman SR, White IR, et al: Propensity score
analysis with partially observed covariates: How should multiple
imputation be used? Statistical methods in medical research 28:3--19,
2019

\bibitem[\citeproctext]{ref-pishgar2020matchthem}
\textbf{43}. Pishgar F, Greifer N, Leyrat C, et al: MatchThem:: Matching
and weighting after multiple imputation. arXiv preprint arXiv:200911772,
2020

\bibitem[\citeproctext]{ref-rubin2018multiple}
\textbf{44}. Rubin DB: Multiple imputation, in Flexible imputation of
missing data, second edition. Chapman; Hall/CRC, 2018, pp 29--62

\bibitem[\citeproctext]{ref-van2011mice}
\textbf{45}. Van Buuren S, Groothuis-Oudshoorn K: Mice: Multivariate
imputation by chained equations in r. Journal of statistical software
45:1--67, 2011
